Information within words in English, French and German speech

### 2019-07-05 surprisal/phoneme_writeup.Rmd
description of histograms and brief general discussion

How is information spread among the sounds in English words? A salient feature of African American English Vernacular is the reduction of final consonants and consonant clusters. Why hasn't this created communication problems, namely identifying and disambiguating words? One possibility is that the information important for identifying a word is not uniformly distributed across all of its characters or sounds, but instead concentrated at the beginnings or middles of words. In these cases, reducing final sounds might reduce the articulatory complexity of words without much cost for the listener in processing these words and inferring their intended meaning. By the time the listener hears only the first sounds of a word, they may already have identified which word they are hearing.

@@

    all_data %>%
    filter(language == select_language, rep == select_rep) %>%
    distinct(word_length, n, rep) %>%
    ggplot(aes(x = word_length, y = n)) +
    geom_col(fill = "white", color = "black") +
    xlab(paste0(str_to_sentence(select_rep), " Length")) +
    ylab("Frequency")
}
```

We use the Providence corpus from CHILDES as data for English. In the plots below, each plot represents the information in each letter, divided by words of different lengths. Word lengths are given by number of letters.

@@

We see a distinctive shape for information at the level of individual letters in English. The first letter has relatively high information, the second letter has less information and the third letter has high infomration, and the final letter has less information (if the word has more than $3$ letters).

@@

@@

For our final analysis, we're going to look at how the information within words in French and German compare to English. We use the Palasis French corpus and Wagner German corpus from CHILDES for our data. We chose French because for many French words the final letters are not pronounced regardless of how the letters in the word appear. For example, "ils sauraient" meaning "they would know" is pronounced without the final "ent" in "sauraient". We are interested in German because German has a much richer case system than English or French, where case markings in German are suffixes at the ends of words. We think that in German, the ends of words may be more important than in French or English.

@@

@@

@@

>>>>

How is information spread among the sounds in English words? A salient feature of African American English Vernacular is the reduction of final consonants and consonant clusters. Why hasn't this created communication problems, namely identifying and disambiguating words? One possibility is that the information important for identifying a word is not uniformly distributed across all of its characters or sounds, but instead concentrated at the beginning or middle of the word. In these cases, reducing final sounds might reduce the articulatory complexity of words without much cost for the listener in processing these words and inferring their intended meaning. By the time the listener hears only the first sounds of a word, they may already have identified which word they are hearing.

@@

    all_data %>%
    filter(language == select_language, rep == select_rep) %>%
    distinct(word_length, n, rep) %>%
    ggplot(aes(x = word_length, y = n)) +
    geom_col(fill = "white", color = "black") +
    xlab("Word length") + 
    ylab("Frequency")
}
    # xlab(paste0(str_to_sentence(select_rep), " Length")) +

```

We use the Providence corpus from CHILDES as data for English. In the plots below, each plot represents the information in each letter, divided by words of different lengths. Word lengths are given by number of letters. See below a histogram of the distribution of word lengths in the Providence corpus. We observe that most words ($84%$) have $5$ letters or fewer, while $99%$ of words in the corpus have fewer than $10$ letters. There are a tiny number of outliers, mainly onomatopoeia, which have $10$ or more letters.

```{r get cumsums for letter and phone frequencies, include = F, eval = F}
get_freq_sums <- function(select_language = "English", select_rep = "letter") {

  all_data %>% 
    filter(language == select_language, rep == select_rep) %>% 
    distinct(word_length, n, rep) %>% 
    mutate(share = n/sum(n)) %>% 
    mutate(cumsum = cumsum(share)) %>% 
    mutate(language = select_language)
}
```

@@

We see a distinctive shape for information at the level of individual letters in English. The first letter has relatively high information, the second letter has less information and the third letter has high information, and the final letter has less information (if the word has more than $3$ letters).

@@

As seen in the histogram below, almost all words in the Providence corpus ($92%$) have four or fewer phones. $99.5%$ have seven or fewer phones.

@@

For our final analysis, we're going to look at how the information within words in French and German compare to English. We use the French Palasis corpus and German Wagner corpus from CHILDES for our data. We chose French because for many French words the final letters are not pronounced regardless of how the letters in the word appear. For example, "ils sauraient" meaning "they would know" is pronounced without the final "ent" in "sauraient". We are interested in German because German has a much richer case system than English or French, where case markings in German are suffixes at the ends of words. We think that in German, the ends of words may be more important than in French or English.

@@

First, we examine the word length frequencies by number of letters in the French Palasis corpus. In the Palasis corpus, $99%$ of words have $10$ letters or fewer. This French corpus has a higher share of words with six to ten letters than the English Providence corpus: $24%$ versus $16%$. However, almost all words in both corpora have $10$ or fewer letters. While there are relatively more medium length words in the French corpus compared to the English corpus, there are very few long words in either corpus.

@@

In the German Wagner corpus, the distribution of word lengths in number of letters has a thicker tail: $99%$ of words have $13$ or fewer letters, compared to the Palasis and Providence corpora where over $99%$ of words had $10$ or fewer letters. However, $97.4%$ of German words have $10$ or fewer letters, so the corpus is similarly comprised of short and medium-length words.

@@

# General Discussion

Across all three languages we examined and both letters and phonemes in English, we found that the first two or three letters and sounds in a word contain relatively more information than any other letters or sounds in the word. In particular, when we knew the letters or sounds that came before, the middles and ends of words became predictable and held little information. Even for a language like German which marks case at the end of nouns, listeners can identify and disambiguate the words they hear by the time they have heard the first few sounds.
